---
title: "B_AURORA_exploring_phenotype_data"
output:
  html_notebook:
    toc: yes
    toc_float: yes
date: "Jan. 21, 2020"
---

```{r}
.libPaths( c( "/opt/R/3.6.2/lib/R/library" , .libPaths() ) )
.libPaths()
```


```{r}
library(tidyverse)

library(readxl)
require(data.table)
```


## load relevant datasets

```{r}
# phenotye and demographinc data
aur_pheno <- read_delim("/data/humgen/guffantilab/AURORA/Aurora_Freeze2_PID/AURORA_Freeze_2_1.0.csv", col_names = TRUE, delim = ",")
# genotyped samples
aur_gen1 <- read_delim("/data/humgen/burook/data_raw/aur/au1.fam", col_names = FALSE, delim = " ")
# a key that contains cross reference between the above two
cross_ref1 <- read_excel("/data/humgen/burook/AURORA/Broad_PID_crossref.xlsx")

# read the second sheet which contains the imputed phenotypes (particularly PDI) as well
pheno_new1 <- read_excel("/data/humgen/burook/AURORA/Bio_PDI_Supplement.xlsx", sheet =2) 

cross_ref2 <- cross_ref1[,c(5,6)]
colnames(cross_ref2) <- c("FID", "PID")

colnames(aur_gen1) <- c("FID","IID","p1","p2","gend","phen")

# combine the pheno and geno data using the cross reference
aur_pheno1 <- cross_ref2 %>% 
  full_join(aur_pheno, by="PID") %>% 
  full_join(aur_gen1, by="FID") %>% 
  full_join(pheno_new1, by="PID") 


write.table(aur_pheno1, "/data/humgen/burook/AURORA/aur_pheno1.txt", row.names = FALSE, col.names = TRUE, quote = FALSE, sep = "\t")

```


```{r}
aur_pheno2 <- aur_pheno1 %>% 
  filter(!is.na(FID)) %>% 
  subset(select=c("FID", "IID", "PID", "ED_PDI_RS.x", "ED_GenderNow", "ED_Age", "BMI", "ED_Marital", "ED_RaceEthCode", "ED_HighestGrade", "M3_PCL5_RS", "M3_PCL5", "M3_PROM_Dep8b_T", "M3_PROM_Dep8b", "WK2_CTQSF_Total_RS", "ED_Event_BroadClass", "ED_Event_SpecificClass")) %>% 
  mutate(EA4 = as.numeric(cut(WK2_CTQSF_Total_RS, breaks = c(0,10,12,14,16), right=T))) %>% 
  print()

```
## PTSD diagnosis and severity score

Total severity score based on a 20 item questionaire (resulting in a range 0 to 80). We used this score taken at 3 months follow-up survey.
```{r}
#pdf("/data/humgen/burook/AURORA/figures/hist_M3_PCL5_RS1.pdf", width=5.5, height=5)
plot1 <- aur_pheno2 %>% 
  filter(!is.na(M3_PCL5_RS)) %>% 
ggplot(aes(M3_PCL5_RS)) + 
  geom_histogram(aes(y =..density..),  
                 col="darkgreen", 
                 fill="navyblue", 
                 alpha=.2, binwidth=0.05) + 
  geom_vline(xintercept=28, linetype="dashed", color = "red") +
  scale_x_log10() +
  geom_density(col=1)
plot1

#dev.off()
#plot1
```

A cut point of 28 (instead of 33 as usual or recommended?) was used to determine a binary diagnosis indicator variable. PTSD cases are those with a score of 28 or more.
```{r}
ggplot(data=aur_pheno2, aes(x=factor(M3_PCL5), y=M3_PCL5_RS)) + 
  geom_boxplot() +
  xlab("PTSD Diagnosis (M3_PCL5)") + ylab("PTSD Severity Score (M3_PCL5_RS)") +
#  xlim("0","1") +       # show only cases and controls (exclude NA's)
  coord_flip()
```

Let's briefly summarize some relevant survey scores for the PTSD cases and controls.
```{r}
aur_pheno2 %>% 
  group_by(M3_PCL5) %>% 
  summarise(
            n=n(),
            mean_PCL = mean(M3_PCL5_RS, na.rm = TRUE),
            mean_CTQSF = mean(WK2_CTQSF_Total_RS, na.rm = TRUE),
            mean_PROM_Dep = mean(M3_PROM_Dep8b_T, na.rm = TRUE)) %>% 
  print()
```
Let's redo the above plot with number of cases and controls inserted
```{r}
#pdf("/data/humgen/burook/AURORA/figures/box_M3_PCL51.pdf", width=5.5, height=5)
plot1 <- aur_pheno2 %>% 
  ggplot(aes(x=factor(M3_PCL5), y=M3_PCL5_RS)) + 
  geom_boxplot() +
  scale_x_discrete(labels=c("Controls (n=461)", "Cases (n=348)", "NA (n=478)")) +
  xlab("PTSD Diagnosis (M3_PCL5)") + ylab("PTSD Severity Score (M3_PCL5_RS)") +
#  xlim("0","1") +       # show only cases and controls (exclude NA's)
  coord_flip()
plot1

#dev.off()
#plot1
```

## MDD diagnosis and severity score
Total severity score based on PROMIS Depression - Short Form 8b (raw score range from 8 to 40). (Note that this raw score is not available.) This raw score is standardized to a T-score with a mean of 50 and standard deviation of 10 (resulting in a range 37.1-81.1). We used this score taken at 3 months follow-up survey.
```{r}
#pdf("/data/humgen/burook/AURORA/figures/hist_M3_PROM_Dep8b_T1.pdf", width=5.5, height=5)
plot1 <- aur_pheno2 %>% 
  ggplot(aes(M3_PROM_Dep8b_T)) + 
  geom_histogram(aes(y =..density..), 
                 col="darkgreen", 
                 fill="navyblue", 
                 alpha=.2, binwidth=1) + 
  geom_vline(xintercept=60, linetype="dashed", color = "red") +
  geom_density(col=1)
plot1

#dev.off()
#plot1
```

A cut point of 60 of the T-score was used to determine a binary diagnosis indicator variable. Depression cases are those with a T-score of 60 or more.
```{r}
ggplot(data=aur_pheno2, aes(x=factor(M3_PROM_Dep8b), y=M3_PROM_Dep8b_T)) +
  geom_boxplot() +
  xlab("MDD Diagnosis (M3_PROM_Dep8b)") + ylab("Depressive Symptoms T-Score (M3_PROM_Dep8b_T)") +
#  xlim("0","1") +       # show only cases and controls (exclude NA's)
  coord_flip()
```

Let's briefly summarize some relevant survey scores for the MDD cases and controls.
```{r}
aur_pheno2 %>% 
  group_by(M3_PROM_Dep8b) %>% 
  summarise(
            n=n(),
            mean_PCL = mean(M3_PCL5_RS, na.rm = TRUE),
            mean_CTQSF = mean(WK2_CTQSF_Total_RS, na.rm = TRUE),
            mean_PROM_Dep = mean(M3_PROM_Dep8b_T, na.rm = TRUE)) %>% 
  print()
```

Let's redo the above plot with number of cases and controls inserted
```{r}
#pdf("/data/humgen/burook/AURORA/figures/box_M3_PROM_Dep8b1.pdf", width=5.5, height=5)
plot1 <- aur_pheno2 %>% 
ggplot(aes(x=factor(M3_PROM_Dep8b), y=M3_PROM_Dep8b_T)) +
  geom_boxplot() +
  scale_x_discrete(labels=c("Controls (n=589)", "Cases (n=226)", "NA (n=472)")) +
  xlab("MDD Diagnosis (M3_PROM_Dep8b)") + ylab("Depressive Symptoms T-Score (M3_PROM_Dep8b_T)") +
#  xlim("0","1") +       # show only cases and controls (exclude NA's)
  coord_flip()
plot1

#dev.off()
#plot1
```

## Childhood trauma

```{r}
ggplot(data=aur_pheno2, aes(WK2_CTQSF_Total_RS)) + 
  geom_histogram(aes(y =..density..),  
                 col="red", 
                 fill="green", 
                 alpha=.2, binwidth=1) + 
  geom_density(col=1) 

```


```{r}
x=table(aur_pheno2[,c("M3_PCL5","EA4")])
y <- x[c(2,1),]
barplot(y)

#par(mfrow=c(1, 1), mar=c(5, 5, 5, 10), xpd=TRUE)
barplot(prop.table(y, margin = 2), beside=F, horiz=F,
        col=c("red","steelblue"),
        legend = rownames(y), legend.text = TRUE,
        args.legend = list(x = "topright", bty = "n", inset=c(-0.25,0)),
        cex.lab=1.5,
        main = "Conditional probability of PTSD diagnosis",
        ylab = "Proportion with PTSD",
        xlab = "Education level",
        names.arg = c("H.S. Diploma \n or less","A.A. Degree \n(2 yrs)","Bachelor's \n Degree (4 yrs)","Master's \n Degree or more"))

```



```{r}
# NB:
# standard errors = standard deviation/square root of number of samples
# i.e., se=sd/sqrt(n)

aur_pheno2 %>% 
  mutate(PTSD_status = factor(M3_PCL5, labels = c("Controls","Cases"))) %>% 
  filter(!is.na(PTSD_status)) %>% 
  group_by(PTSD_status) %>% 
  summarise(n=n(),
            # PCL score
            PCL_mean = mean(M3_PCL5_RS, na.rm = TRUE),
            PCL_se = sd(M3_PCL5_RS, na.rm = TRUE)/sqrt(n()),
            #Depressive symptom score
            Dep_mean = mean(M3_PROM_Dep8b_T, na.rm = TRUE),
            Dep_se = sd(M3_PROM_Dep8b_T, na.rm = TRUE)/sqrt(n()),
            #Age
            Age_mean = mean(ED_Age, na.rm = TRUE),
            Age_se = sd(ED_Age, na.rm = TRUE)/sqrt(n()),
            #BMI
            BMI_mean = mean(BMI, na.rm = TRUE),
            BMI_se = sd(BMI, na.rm = TRUE)/sqrt(n()),
            #Childhood trauma score
            CT_mean = mean(WK2_CTQSF_Total_RS, na.rm = TRUE),
            CT_se = sd(WK2_CTQSF_Total_RS, na.rm = TRUE)/sqrt(n())) %>% 
  print() 

```


```{r}
# install necessary packages
#install.packages("arsenal",  method = "curl", lib = "/PHShome/bm363/R/x86_64-pc-linux-gnu-library/3.6")
#install.packages("Amisc",  method = "curl", lib = "/PHShome/bm363/R/x86_64-pc-linux-gnu-library/3.6")
#install.packages("table1",  method = "curl", lib = "/PHShome/bm363/R/x86_64-pc-linux-gnu-library/3.6")
# we can use this package if we need significance tests conducted
library(arsenal)
sum_tab1 <- tableby(M3_PCL5 ~ M3_PCL5_RS + M3_PROM_Dep8b_T + ED_Age + as.factor(ED_GenderNow) + BMI + ED_HighestGrade +  WK2_CTQSF_Total_RS, data=aur_pheno2, test=T);
summary(sum_tab1)

# 1=Hispanic, 2=White, 3=Black, 4=Asian, and 5=Asian/other
# 1: Male; 2: Female; 3: Transgender; 4: Does not identify

# a better lookong descriptive tables can be generated with the following package
library(table1)
tab_data1 <- aur_pheno2 %>% 
  mutate(PTSD_status = factor(M3_PCL5, labels = c("Controls","Cases"))) %>% 
  mutate(ED_GenderNow = factor(ED_GenderNow, labels = c("Male","Female", "Transgender"))) %>% 
  mutate(Gender = factor(ED_GenderNow, labels = c("Male","Female", "Transgender"))) %>% 
  mutate(ED_RaceEthCode = factor(ED_RaceEthCode, labels = c("Hispanic", "White", "Black", "Asian"))) %>% 
  filter(!is.na(PTSD_status))

label(tab_data1$M3_PCL5_RS) <- "PCL score"
label(tab_data1$M3_PROM_Dep8b_T) <- "Depressive symptom T score"
label(tab_data1$ED_Age) <- "Age"
label(tab_data1$ED_GenderNow) <- "Gender"
label(tab_data1$ED_RaceEthCode) <- "Race/Ethnicity"
label(tab_data1$WK2_CTQSF_Total_RS) <- "Childhood trauma score"

sum_tab2 <- table1(~ M3_PCL5_RS + M3_PROM_Dep8b_T + ED_Age + ED_GenderNow + BMI + ED_HighestGrade +  WK2_CTQSF_Total_RS + ED_RaceEthCode | PTSD_status, 
                   data = tab_data1)#, overall = FALSE)
sum_tab2


```

Descriptive stat table with MDD case and control

```{r}

# 1=Hispanic, 2=White, 3=Black, 4=Asian, and 5=Asian/other
# 1: Male; 2: Female; 3: Transgender; 4: Does not identify

# a better lookong descriptive tables can be generated with the following package
library(table1)
tab_data1 <- aur_pheno2 %>% 
  mutate(MDD_status = factor(M3_PROM_Dep8b, labels = c("Controls","Cases"))) %>% 
  mutate(ED_GenderNow = factor(ED_GenderNow, labels = c("Male","Female", "Transgender"))) %>% 
  mutate(Gender = factor(ED_GenderNow, labels = c("Male","Female", "Transgender"))) %>% 
  mutate(ED_RaceEthCode = factor(ED_RaceEthCode, labels = c("Hispanic", "White", "Black", "Asian"))) %>% 
  filter(!is.na(MDD_status))

label(tab_data1$M3_PCL5_RS) <- "PCL score"
label(tab_data1$M3_PROM_Dep8b_T) <- "Depressive symptom T score"
label(tab_data1$ED_Age) <- "Age"
label(tab_data1$ED_GenderNow) <- "Gender"
label(tab_data1$ED_RaceEthCode) <- "Race/Ethnicity"
label(tab_data1$WK2_CTQSF_Total_RS) <- "Childhood trauma score"

sum_tab2 <- table1(~ M3_PCL5_RS + M3_PROM_Dep8b_T + ED_Age + ED_GenderNow + BMI + ED_HighestGrade +  WK2_CTQSF_Total_RS + ED_RaceEthCode | MDD_status, 
                   data = tab_data1)#, overall = FALSE)
sum_tab2

```

```{r}
# The PTSD diagnosis (M3_PCL5) is based on the score (M3_PCL5_RS). 
# But we have samples with diagnosis but not score.
aur_pheno2 %>% 
  subset(select=c("M3_PCL5","M3_PCL5_RS")) %>% 
  filter(is.na(M3_PCL5_RS)) %>% 
  filter(!is.na(M3_PCL5)) %>% 
  print()
```

Joint genotype IID and with phenotype label.

```{r}
aur_pheno3 <- cross_ref2 %>% 
  full_join(pheno_new1, by="PID") %>% 
  full_join(aur_gen1, by="FID") 

aur_pheno3 %>% 
  filter(!is.na(FID)) %>% 
  group_by(ED_PDI) %>% 
  summarise(
    n=n(),
#    mean_PCL = mean(ED_PDI_RS, na.rm = TRUE),
    ) %>%
  print()
```

Extract and save phenotype file for plink (to use with --make-pheno).

```{r}
tmp3<-aur_pheno3 %>% 
  filter(!is.na(FID)) %>% 
  subset(select=c("FID","IID","ED_PDI")) %>% 
  print()

#write.table(tmp3, "/data/humgen/burook/AURORA/ED_PDI.txt", row.names = FALSE, col.names = TRUE, quote = FALSE, sep = "\t")

```


```{r}

aur_pheno2 %>% 
  filter(!is.na(FID)) %>%   # consider only genotyped samples
  group_by(M3_PCL5) %>%
  summarise(n=n()) %>%
  mutate(n_perc = 100*n / sum(n)) %>% 
  print()

aur_pheno2 %>% 
  filter(!is.na(FID)) %>% 
  group_by(M3_PCL5) %>% 
  summarise(
    n=n(),
    mean_PCL = mean(M3_PCL5_RS, na.rm = TRUE),
    mean_Dep8b_T = mean(M3_PROM_Dep8b_T, na.rm = TRUE)
    ) %>%
  print()

aur_pheno2 %>% 
  filter(!is.na(FID)) %>% 
  group_by(M3_PCL5) %>% 
  summarise(
    q = list(quantile(M3_PCL5_RS, na.rm = TRUE))
    ) %>%
  unnest() %>% 
  print()

```



